---
title: "Visualizing Predictions"
author: Kris Sankaran
date: 2026-09-01
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
    fig-align: center
execute:
    echo: true
    message: false
    warning: false
    cache: false
    eval: true
---

[Reading](https://ema.drwhy.ai/partialDependenceProfiles.html)
[Code]()

```{r}
#| label: setup
#| echo: false
library(knitr)
library(tidyverse)
theme_set(theme_classic())
set.seed(2026)
```

## Motivation

1. Partial dependence profiles (PDP), individual conditional expectation (ICE)
plots, and accumulated local effects (ALE) are types of global explainability
methods. They can be applied post-hoc to any model and try to reveal how the
model relates a predictor variable/set of predictor variables with the
response.

2. Partial dependence profiles were first introduced as a minor point in an
influential paper about gradient boosting. ICE and ALE were introduced to
address some limitations of PDP, mainly related to the fact that prediction
surfaces can become complicated in high-dimensions and when many interactions
are present.

## Partial Dependence Profiles

3. Let $f$ be the learned black box function. Let $x$ be its $D$-dimensional
input. The idea is to ask how $f$ changes as we vary the $d$th coordinate over
its range. We need to do something with the remaining coordinates; the idea of
PDP is to average across them.

$$
\text{PDP}_d(z) = \mathbb{E}_{p(x_{-d})}\left[f(x_d = z, x_{-d})\right] \approx \frac{1}{N}\sum_{n=1}^N f(x_d = z, x_{n, -d})
$$

The highlighted part refers to manually intervening on the $d$th coordinate and
setting it to $z$. If we evaluate this average across many different values of
$z$, we get a PDP curve. Note that for each value $z$, we compute an average
across the entire dataset -- this can get slow.

4. This definition generalizes to larger-dimensional subsets of variables, but
we'll stick to one-dimensional curves for simplicity.

5. Here is some geometric intuition. We take the original samples, manually set
all their $d$th coordinates to $z$, evaluate the function $f$, and then
average. We repeat this for many $z$'s to trace out the partial dependence
curve.

![Evaluating the PDP at a single value $z = x_d^*$ (top) and repeating this
across many $z$ to trace out the $\text{PDP}_d(x_d)$ curve
(bottom).](figures/pdp_slice_sketch.png)

## Individual Conditional Expectations

6. ICE plots consider each term in the PDP formula as its own function of $z$.

$$
h_n(z) := f(x_d = z, x_{n, -d})
$$

This has the nice "Ceteris Paribus" interpretation. For each sample, it traces
out the curve where the $d$th coordinate is set to different values $z$ but all
other coordinates are held fixed.

7. The key insight is that these curves can be meaningful on their own. The
averaging step in PDP can mask some meaningful variation in model predictions.
This is especially useful for discovering interactions. For example, here are
ICE curves for a model trained to predict survival with the Titanic dataset.
The age effect is very different for crew vs. passengers.

![ICE ("Ceteris Paribus") curves for a survival model fit to the Titanic
dataset, colored by passenger/crew class.](figures/plot_titanic_ice.png)

8. Two common tricks for ICE plots:

* We might suspect an interaction with an unmeasured variable. This can
sometimes be seen as groups of curves with different shapes, but where we
don't have a variable in the data that splits those groups. In this case,
people often cluster the curves and use cluster membership as a stand in for
the unobserved interacting variable.

![Groups of ICE curves with different shapes can suggest an interaction with
an unmeasured variable; clustering the curves gives a stand-in for that
variable.](figures/ice_cluster_sketch.png)

* Often the curves are offset because of differences in the coordinates other
than the $d$th. To see whether they nonetheless have similar shapes, it's
sometimes helpful to "center" the curves, so that they all start out equal to
one another.

$$
\tilde{h}_n(z) = f(x_d = z, x_{n, -d}) - f(x_d = 0, x_{n, -d})
$$

![Raw ICE curves for a partial $\hat{y}$ vs. age relationship (left) and the
same curves after centering (right); centering makes it easier to compare
shapes across curves that start at different levels.](figures/plot_partial_yhat.png)

## Effects and Partial Residual Plots

9. Fox and Weisberg (2018) focus on visualizing effects just in the case of
linear regression models. Even here, it pays to be careful when thinking about
the role of interactions.

10. Suppose we are interested in the $d$th predictor in a linear regression
model. They partition the predictors into $(z, v, u)$:

* $z$ - the predictor we are interested in.
* $v$ - the predictors that interact with $z$.
* $u$ - other unrelated predictors.

11. They recommend identifying a grid of values across $v$ and then tracing the
function of $z$ with $v$ fixed to each of those grid points.

$$
\left\{h_d(z, v)\right\}_{v \in \mathcal{G}} \equiv \left\{\begin{pmatrix}z \\ v \\ 0\end{pmatrix}^T\hat\beta\right\}_{v \in \mathcal{G}}
$$

In this example $z$ = per person GDP and $v$ = country group.

![The ppgdp predictor effect plot: fitted infant mortality vs. per-capita GDP,
traced separately for each of a grid of country groups
(oecd/other/africa).](figures/plot_ppgdp_effect.png)

12. They also suggest overlying the residuals on these effects plots.

$$
\begin{pmatrix}z = x_{nd} \\ v\end{pmatrix}^T \hat\beta + e(x_n)
$$

(I don't like their notation for this... though I don't know if this is any
better.)

Regardless, the idea is simple: This plot can help identify systematic
structure in the residuals. E.g., if the residuals are systematically
different in some regions of the $z$-space, then we might have a
mis-specified model. For example, in this toy example, the residuals suggest
that the model is missing an interaction term between x2 and x3.

![The x2*x3 effect plot with residuals overlaid; the trend in the residuals
across facets suggests a missing interaction between x2 and x3.](figures/plot_x2x3_effect.png)

## Probabilistic Models

13. So far, we've been focusing on visualizign black box models. In many
settings, we might instead want to fit probabilistic models directly informed by
our conceptual understanding.  Visualization fits nicely into this workflow as
well, and can help in building up more realistic, but still interpretable,
models.

14. Exploratory data analysis and model building work synergistically. In
real-world applications, visualization guides selection toward more reasonable
models.

15. Models' exact forms are typically unknown initially, though several
plausible candidates usually exist. Exploratory work eliminates some options
while revealing unexpected connections.

16. As an illustration, consider PM2.5 monitoring. A researcher is interested in
monitoring the level of PM2.5, a type of small air particulate that can be bad
for public health. High-quality weather station data apply locally; lower
quality satellite data provide global coverage. A calibration model using
station measurements to adjust satellite readings is needed.

```{r}
#| label: pm25-data
f <- tempfile()
download.file("https://github.com/jgabry/bayes-vis-paper/blob/master/bayes-vis.RData?raw=true", f)
GM <- get(load(f))
GM@data <- GM@data |>
    mutate(
        log_pm25 = log(pm25),
        log_sat = log(sat_2014)
    )
```

17. The baseline model uses the equation $\text{station} = a + b \times
\text{satellite}$ at co-located measurement points. The Global Burden of
Disease project employed this approach through 2016.

```{r}
#| label: pm25-baseline
ggplot(GM@data, aes(log_sat, log_pm25)) +
    geom_point(aes(col = super_region_name), size = 0.8, alpha = 0.7) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "log(satellite)", y = "log(ground station)", col = "WHO Region") +
    coord_fixed()
```

18. Plotting reveals substantial heterogeneity and large residuals. The key
question: What factors correlate with residuals that could enhance model
performance?

* Error terms $\epsilon_i$ in equations $y_i = f(x_i) + \epsilon_i$ represent
unmeasured factors determining $x$-$y$ relationships.
* Desert sand increases PM2.5 but remains invisible to satellites; residuals
likely correlate with desert locations (PM2.5 underpredicted), suggesting
inclusion of geographic features improves fit.

19. One hypothesis: country region significance. Fitting separate regression
lines per WHO super-region reveals slope variation, suggesting models should
incorporate region-specific slopes.[^interaction]

```{r}
#| label: pm25-regions
ggplot(GM@data, aes(log_sat, log_pm25)) +
    geom_point(aes(col = super_region_name), size = 0.4, alpha = 0.7) +
    geom_smooth(aes(col = super_region_name), method = "lm", se = FALSE, size = 2) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "log(satellite)", y = "log(ground station)", col = "WHO Region") +
    coord_fixed()
```

20. WHO categorizations are somewhat arbitrary. An alternative approach is to
cluster ground stations by PM2.5 level to create problem-specific regional
groupings.

```{r}
#| label: pm25-clusters
average <- GM@data |>
    group_by(iso3) |>
    summarise(pm25 = mean(pm25))

clust <- dist(average$pm25) |>
    hclust() |>
    cutree(k = 6)

GM@data$cluster_region <- as.character(clust[match(GM@data$iso3, average$iso3)])
ggplot(GM@data, aes(log_sat, log_pm25)) +
    geom_point(aes(col = cluster_region), size = 0.4, alpha = 0.7) +
    geom_smooth(aes(col = cluster_region), method = "lm", se = FALSE, size = 2) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "log(satellite)", y = "log(ground station)", col = "Cluster Region") +
    coord_fixed()
```

21. A model network emerges; refined comparison tools become necessary. This
sets up the subject of our next notes: prior and posterior predictive checks.

[^interaction]: Alternatively, this can be viewed as introducing an
interaction between satellite measurements and WHO region.

## Prior and Posterior Predictives

22. How should we go about choosing between the plausible models developed from
exploratory analysis? An alternative which often brings more insight into the
structure of the problem than raw test-set error is to use prior and posterior
predictive distributions for (visual) model comparison.

```{r}
#| label: predictive-setup
library(rstan)
```

23. The prior predictive distribution helps decide whether certain model
families are reasonable candidates for a problem, before formally
incorporating the evidence coming from the data. The mechanism is: (a) sample
parameters from the prior, and (b) simulate data from the model with those
parameters.

24. Vague priors can produce implausible simulated datasets, despite being
"more objective" in some sense.

```{r}
#| label: prior1
prior1 <- function(Nsim) {
    tau0 <- 1 / sqrt(rgamma(1, 1, rate = 100))
    tau1 <- 1 / sqrt(rgamma(1, 1, rate = 100))
    sigma <- 1 / sqrt(rgamma(1, 1, rate = 100))
    beta0i <- rnorm(7, 0, tau0)
    beta1i <- rnorm(7, 0, tau1)
    beta0 <- rnorm(1, 0, 100)
    beta1 <- rnorm(1, 0, 100)

    epsilon <- rnorm(Nsim, 0, sigma)
    data.frame(
        log_pm25 = GM$log_pm25,
        region = GM$super_region_name,
        sim = beta0 + beta0i[GM$super_region] + (beta1 + beta1i[GM$super_region]) * GM$log_sat + epsilon
    )
}

prior1_data <- map_dfr(1:12, ~ prior1(Nsim = nrow(GM@data)), .id = "replicate")
ggplot(prior1_data, aes(log_pm25, sim)) +
    geom_abline(slope = 1) +
    geom_point(aes(col = region), alpha = 0.1, size = 0.4) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "True station data", y = "Simulated station data") +
    facet_wrap(~replicate, scale = "free_y")
```

Prior predictive samples from this vague prior are on a completely implausible
scale.

25. Subjective, informative priors instead produce more plausible samples,
lying in a comparable range to the true data. Note, however, that samples from
the prior predictive do not need to look exactly like the observed data -- if
they did, there would be no need to fit model parameters!

```{r}
#| label: prior2
prior2 <- function(Nsim) {
    tau0 <- abs(rnorm(1, 0, 1))
    tau1 <- abs(rnorm(1, 0, 1))
    sigma <- abs(rnorm(1, 0, 1))
    beta0i <- rnorm(7, 0, tau0)
    beta1i <- rnorm(7, 0, tau1)
    beta0 <- rnorm(1, 0, 1)
    beta1 <- rnorm(1, 1, 1)

    epsilon <- rnorm(Nsim, 0, sigma)
    data.frame(
        log_pm25 = GM$log_pm25,
        region = GM$super_region_name,
        sim = beta0 + beta0i[GM$super_region] + (beta1 + beta1i[GM$super_region]) * GM$log_sat + epsilon
    )
}

prior2_data <- map_dfr(1:12, ~ prior2(Nsim = nrow(GM@data)), .id = "replicate")
ggplot(prior2_data, aes(log_pm25, sim)) +
    geom_abline(slope = 1) +
    geom_point(aes(col = region), alpha = 0.1, size = 0.4) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "True station data", y = "Simulated station data") +
    facet_wrap(~replicate, scale = "free_y")
```

26. Philosophically, this prior predictive analysis is based on the idea that,
though probability is subjective, evidence can be used to update our beliefs.
The idea of the prior predictive is to visually encode subjective beliefs
about the problem under study before gathering new evidence.

27. After calibrating the prior predictive, models are fit. The posterior
predictive is like the prior predictive, except that it samples model
parameters from the data-informed posterior, rather than the data-ignorant
prior: it represents the distribution of new datasets when drawing parameters
from the posterior. The simulation mechanism is (a) draw model parameters from
the posterior and (b) simulate a dataset using parameters from (a).

```{r}
#| label: fit-models
datasets <- list(
    "lm" = with(GM@data, list(N = length(log_pm25), y = log_pm25, x = log_sat)),
    "regions" = with(
        GM@data,
        list(N = length(log_pm25), y = log_pm25, x = log_sat, group = super_region, R = n_distinct(super_region))
    ),
    "clusters" = with(
        GM@data,
        list(N = length(log_pm25), y = log_pm25, x = log_sat, group = as.integer(cluster_region), R = n_distinct(cluster_region))
    )
)

model_def <- list(
    "lm" = stan_model("https://uwmadison.box.com/shared/static/hoq1whain301bj6gj670itxabnnhvcy7.stan"),
    "hier" = stan_model("https://uwmadison.box.com/shared/static/lvouz9jj4rbkmrx5osj2dtrhj2ycdll8.stan")
)

controls <- list(max_treedepth = 15, adapt_delta = 0.99)
models <- list(
    "lm" = sampling(model_def$lm, data = datasets$lm, chains = 1, control = controls),
    "regions" = sampling(model_def$hier, data = datasets$regions, chains = 1, control = controls),
    "clusters" = sampling(model_def$hier, data = datasets$clusters, chains = 1, control = controls)
)
```

We've defined three models here: an ordinary linear regression (`lm`), a
region-based hierarchical model, and a cluster-based hierarchical model. The
hierarchical models allow different slopes for different groups.

```{r}
#| label: posterior-predictive
posterior_samples <- as.matrix(models$clusters, pars = "y_sim")[950:961, ] |>
    t() |>
    as_tibble() |>
    bind_cols(GM@data) |>
    pivot_longer(V1:V12, names_to = "replicate", values_to = "log_pm25_sim")

ggplot(posterior_samples, aes(log_pm25, log_pm25_sim)) +
    geom_abline(slope = 1) +
    geom_point(aes(col = cluster_region), size = 0.4, alpha = 0.1) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "True station data", y = "Simulated station data") +
    facet_wrap(~replicate, scale = "free_y")
```

Compared to the prior predictive, the posterior predictive is much more
closely related to the true underlying dataset.

28. One subtlety is that there is a danger of overfitting features in the
posterior predictive. It is best to choose features of the data that are not
directly modeled (e.g., if you use slope in the model estimation, then don't
evaluate the posterior predictive using the slope, since by definition this
will be well-captured).

```{r}
#| label: skewness-comparison
apply_stat <- function(x, f) {
    z <- as.matrix(x, pars = "y_sim")
    tibble(replicate = seq_len(nrow(z)), statistic = apply(z, 1, f))
}

skew <- function(x) {
    xdev <- x - mean(x)
    n <- length(x)
    r <- sum(xdev^3) / sum(xdev^2)^1.5
    r * sqrt(n) * (1 - 1 / n)^1.5
}

posteriors <- map_dfr(models, ~ apply_stat(., skew), .id = "model")
truth <- skew(GM@data$log_pm25)
ggplot(posteriors, aes(statistic)) +
    geom_histogram(aes(fill = model), binwidth = 0.01) +
    geom_vline(xintercept = truth, col = "red") +
    scale_fill_brewer(palette = "Set3")
```

It seems that the model that uses clustering to define regions is able to
simulate datasets with skewness similar to that in the real dataset.
